Introducing statistics with \\texttt{intRo}

### 2014-12-05 student-paper-competition-2014/paper.Rmd
small chages to introduction

The release of \texttt{RStudio}'s \texttt{Shiny}, which allows for easily turning an R-based analysis into an interactive web application, led to the creation of an introductory statistics application which we call \texttt{intRo}. \texttt{intRo} offers a number of key advantages over traditional statistics software, which will be discussed at length in this text.

# What is \texttt{intRo}?
\texttt{intRo} is a web-based tool to accompany and introductory statistics class. It is meant to assist in the learning of statistics rather than a stand-alone deliverer of statistics education, with the intention of being used in conjunction with a guided class.

We have multiple philosophies that guided the creation of \texttt{intRo}, and this comes out in the design. Primarily, \texttt{intRo} is \emph{easy} to use and can be an \emph{exciting} part of learning statistics. Additionally, \texttt{intRo} is an \emph{extensible} tool, allowing for a user of \texttt{intRo} to tailor the tool for his or her own classroom needs.

@@

For example, an instructor may wish to introduce hypothesis testing from a randomization framework rather than distributional. In this case, a non-parametric test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork \texttt{intRo} from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `observe.R`, `output.R`, `reactive.R`, `static.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a \texttt{shiny} application. The `ui.R` file contains all the user interface elements within the new module's tab. This will include any options in the options panel as well as results. The `helper.R` and `static.R` files refer to static functions and global variable that may be used within the module. The file `output.R` contains the reactive elements that tie output to the user interface within \texttt{shiny} and `observe.R` and `reactive.R` contain elements within an `observer` function or `reactive` function within the \texttt{shiny} framework. By placing each of these files within the `modules/nonparametric_test` folder and adding a single line to a navigation file indicating under which module grouping the non-parametric test module should go, a new module is available within \texttt{intRo} upon running the application.

# In the classroom
\texttt{intRo}'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as JMP and Excel, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. JMP, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using \texttt{R} itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like \texttt{rmarkdown} [@rmarkdown], but this again is a bit of a learning curve for the new statistics student).

@@

Reactive programming is a programming paradigm that ``tackles issues posed by event-driven applications by providing abstractions to express programs as reactions to external events and having the language automatically manage the flow of time (by conceptually supporting simultaneity), and data and computation dependencies.'' [@bainomugisha2012survey] As implemented by \texttt{shiny} [@shiny], results automatically update when users interact with the interface.

\texttt{intRo} leverages reactive programming techniques as data is manipulated by the user, this carries throughout the entire session. In a typical \texttt{shiny} app, users interact with inputs, which in turn yield different results. Within \texttt{intRo}, the users are able to interact and manipulate the data underlying the entire application. This posed many challenges in the creation of \texttt{intRo}, namely timely save points according to the user's timeline, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}. \ak{[Maybe add a figure here?, including example code for dynamically loading the ui]} However, \texttt{shiny} does have limits in its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{shiny} allows for arbitrary function calls to to generate these steps, however they must be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{shiny} to limit functionality in favor of usability of their package.

>>>>

Upon the release of \texttt{RStudio}'s \texttt{Shiny} [@shiny] it became easy for an R-based analysis to be converted to an interactive web application. This in turn, led to the creation of an introductory statistics application which we call \texttt{intRo}.

\texttt{intRo} offers a number of key advantages over traditional statistics software, which will be discussed at length in this text.

# What is \texttt{intRo}?
\texttt{intRo} is a web-based tool to accompany an introductory statistics class. It is meant to assist in the learning of statistics rather than a stand-alone deliverer of statistics education, with the intention of being used in conjunction with a guided class.

We have three fundamental philosophies that guided the creation of \texttt{intRo}, and this comes out in the design. Primarily, \texttt{intRo} is \emph{easy} to use and can be an \emph{exciting} part of learning statistics. Additionally, \texttt{intRo} is an \emph{extensible} tool, allowing for a user of \texttt{intRo} to tailor the tool for his or her own classroom needs.

@@

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a randomization framework rather than distributional. In this case, a non-parametric test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork \texttt{intRo} from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `observe.R`, `output.R`, `reactive.R`, `static.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a \texttt{Shiny} application. The `ui.R` file contains all the user interface elements within the new module's tab. This will include any options in the options panel as well as results. The `helper.R` and `static.R` files refer to static functions and global variable that may be used within the module. The file `output.R` contains the reactive elements that tie output to the user interface within \texttt{Shiny} and `observe.R` and `reactive.R` contain elements within an `observer` function or `reactive` function within the \texttt{Shiny} framework. By placing each of these files within the `modules/nonparametric_test` folder and adding a single line to a navigation file indicating under which module grouping the non-parametric test module should go, a new module is available within \texttt{intRo} upon running the application.

# In the classroom
\texttt{intRo}'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as \texttt{JMP} and \texttt{Excel}, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. \texttt{JMP}, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using \texttt{R} itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like \texttt{rmarkdown}, but this again is a bit of a learning curve for the new statistics student).

@@

Reactive programming is a programming paradigm that ``tackles issues posed by event-driven applications by providing abstractions to express programs as reactions to external events and having the language automatically manage the flow of time (by conceptually supporting simultaneity), and data and computation dependencies.'' [@bainomugisha2012survey] As implemented by \texttt{Shiny}, results automatically update when users interact with the interface.

\texttt{intRo} leverages reactive programming techniques as data is manipulated by the user, this carries throughout the entire session. In a typical \texttt{Shiny} app, users interact with inputs, which in turn yield different results. Within \texttt{intRo}, the users are able to interact and manipulate the data underlying the entire application. This posed many challenges in the creation of \texttt{intRo}, namely timely save points according to the user's timeline, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{Shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{Shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}. \ak{[Maybe add a figure here?, including example code for dynamically loading the ui]} However, \texttt{Shiny} does have limits in its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{Shiny} allows for arbitrary function calls to to generate these steps, however they must be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{Shiny} to limit functionality in favor of usability of their package.
